---
title: "Untitled"
author: "bozhidar"
date: "2023-01-08"
output: html_document
---
 loading libraries

```{r}
BiocManager::install("tximport")
BiocManager::install("readr")
BiocManager::install("tximportData")
BiocManager::install("airway")
BiocManager::install("tidyverse")
BiocManager::install("rio")
library("rio")
library("DESeq2")
library("tximport")
library("readr")
library("tximportData")
library("airway")
library("tidyverse")
library("dplyr")
```
importing data


```{r}
txi <- readRDS("D:GSE161860_txi.RDS")
```

 preparing counts data

 merging the data from  GSE161860 with SraRunTable and chosing the part we need``


```{r}
counts_data <- txi[["counts"]]
coldata <-read.csv("D:SraRunTable.txt")
coldata <- coldata[, c("Run", "Treatment")]
```

making sure that row names in coldata matches to column names in counts_data


```{r}
all(colnames(counts_data) %in% coldata$Run)
```

 making sure they are in the same order


```{r}
coldata <- coldata[order(coldata$Run),]
all(colnames(counts_data) == coldata$Run)
```

rename column name


```{r}
coldata <-coldata %>%
  rename_at("Run", ~ "Sample name")
```

 construct a DESeqDataSet object
```{r}


txi$counts <- txi$counts[,coldata$`Sample name`]
txi$abundance <- txi$abundance[,coldata$`Sample name`]
txi$length <- txi$length[,coldata$`Sample name`]
ddsTxi <- DESeqDataSetFromTximport(txi,
                                   colData = coldata, 
                                   design = ~ Treatment)
```




### removing rows with low gene counts and keeping rows that have at least 10 reads total

`` helps reduce the size of the object which increases the computational speed``


```{r}
keep <- rowSums(counts(ddsTxi)) >= 10
ddsTxi <- ddsTxi[keep, ]
```

 running DESeq
 
if needed collapsing the technical replicates``


```{r}
ddsTxi <- DESeq(ddsTxi)
```

saving the results``


```{r}
results_ddsTxi <- results(ddsTxi)
results_ddsTxi
```

 Exploring results


```{r}
summary(results_ddsTxi)
```

contrasts


```{r}
resultsNames(ddsTxi)
```

 MA plot


```{r}
plotMA(results_ddsTxi)
```

ordering results

ordering the results to see the 10 genes with the most significant p value``


```{r}
resOrdered <- results_ddsTxi[order(results_ddsTxi$pvalue),]
head(resOrdered,n=10)
```

reqired libraries

```{r}
BiocManager::install("biomaRt")
BiocManager::install("annotables")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("EnsDb.Hsapiens.v86")
library("biomaRt")
library("org.Hs.eg.db")
library("EnsDb.Hsapiens.v86")
library("tidyverse")
```

input list of Enselbl IDs

```{r}
ensembl.ids <- read.delim("D:Ensembl IDss.txt", header = F)
```

 translating with biomart

seeing the available databeses and their values``


```{r}
listEnsembl()
```

choosing the database``


```{r}
ensembl <- useEnsembl(biomart = "genes")
```

seeing which dataset to use``


```{r}
datasets <- listDatasets(ensembl)
```

connecting the dataset and databese we want to use``


```{r}
ensembl.conection <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
```

seeing which attributes we need``
```{r}
attributes <- listAttributes(ensembl.conection)
```



seeing which filter we need``


```{r}
filters <- listFilters(ensembl.conection)
```

buildung the query``



```{r}
getBM(attributes = c("ensembl_gene_id","external_gene_name"),
      filters = "ensembl_gene_id",
      values = ensembl.ids$V1,
      mart = ensembl.conection)
```

###GSVA



```{r}
BiocManager::install("shinydashboard")
BiocManager::install("shinyjs")
BiocManager::install("limma")
library("shinydashboard")
library("shinyjs")
library("limma")
```


```{r}
rm(list=ls())
 if (!require(GSVA)) BiocManager::install("GSVA")
 if (!require(rio)) install.packages('rio')
 if (!require(here)) install.packages('here')
 if (!require(stringr)) install.packages('stringr')
 if (!require(GSVAdata)) install.packages('GSVAdata')
library(GSVA)
library(rio)
library(here)
library(stringr)
library(org.Hs.eg.db)
```
loading gene sets from GMT file and package (for example comparison)
```{r}
#Take a look at the format of an example set
library(GSEABase)
library(GSVAdata)
library(purrr)
data(c2BroadSets)
class(c2BroadSets)
#Load a gene set we have
gene_sets <- read.delim(here::here("datasmall","h.all.v2023.1.Hs.symbols.gmt"),
sep = "\t", header = FALSE, stringsAsFactors = FALSE)
#Somehow, the format of this gene set seems to require a lot of cleaning
class(gene_sets)
```

Transforming gene set
```{r}
#Transforming gene set rownames
gene_sets$V1<-str_replace(gene_sets$V1,"HALLMARK_","")
rownames(gene_sets)<-gene_sets$V1
#Removing metadata to make analysis easier and straightforward
gene_sets<-gene_sets[,-1:-2]
# Making a list with the desired dimensions. Transposing x and y transforms to matrix
gene_sets<-as.list(as.data.frame(t(gene_sets))) 
#However, the format of gene identifiers should also be changed from symbol to enrezID
hs <- org.Hs.eg.db
gene_sets<-sapply(gene_sets, function(a) select(hs, keys = a,
       columns = c("ENTREZID", "SYMBOL"),
       keytype = "SYMBOL"))
# Manipulation of the resulting list to make it look as desired
gene_sets<-t(gene_sets)
gene_sets<-split(gene_sets, rownames(gene_sets))
gene_sets<-purrr::transpose(gene_sets)
names(gene_sets)<-c("Symbol", "geneIds")
```

Loading expression set data and necessary transformations
```{r}
#Loading an expression set
if (!require(GSVAdata)) BiocManager::install("GSVAdata")
library(GSVAdata)
data(leukemia)
# Exploring the loaded data
class(leukemia_eset)
# Extracting the expression matrix as I had difficulty renaming the row/feature/gene names
expr.matr<-leukemia_eset@assayData[["exprs"]]
rownames(expr.matr)<-str_replace_all(rownames(expr.matr),"[^0-9.-]","")
# Observing the result
head(rownames(expr.matr))
```

Performing the GSVA with the obtained objects
```{r}
leukemia_es <- gsva(expr.matr, gene_sets[["geneIds"]],
            min.sz=10, max.sz=500, verbose=TRUE)
```


```{r}
if (!require(shinydashboard)) BiocManager::install("shinydashboard")
if (!require(shinyjs)) BiocManager::install("shinyjs")
if (!require(limma)) BiocManager::install("limma")
library("shinydashboard")
library("shinyjs")
library("limma")
```

