---
title: "Untitled"
author: "bozhidar"
date: "2023-01-08"
output: html_document
---
 loading libraries

```{r}
BiocManager::install("tximport")
BiocManager::install("readr")
BiocManager::install("tximportData")
BiocManager::install("airway")
BiocManager::install("tidyverse")
BiocManager::install("rio")
library("rio")
library("DESeq2")
library("tximport")
library("readr")
library("tximportData")
library("airway")
library("tidyverse")
library("dplyr")
```
importing data


```{r}
txi <- readRDS("D:GSE161860_txi.RDS")
```

 preparing counts data

 merging the data from  GSE161860 with SraRunTable and chosing the part we need``


```{r}
counts_data <- txi[["counts"]]
coldata <-read.csv("D:SraRunTable.txt")
coldata <- coldata[, c("Run", "Treatment")]
```

making sure that row names in coldata matches to column names in counts_data


```{r}
all(colnames(counts_data) %in% coldata$Run)
```

 making sure they are in the same order


```{r}
coldata <- coldata[order(coldata$Run),]
all(colnames(counts_data) == coldata$Run)
```

rename column name


```{r}
coldata <-coldata %>%
  rename_at("Run", ~ "Sample name")
```

 construct a DESeqDataSet object
```{r}


txi$counts <- txi$counts[,coldata$`Sample name`]
txi$abundance <- txi$abundance[,coldata$`Sample name`]
txi$length <- txi$length[,coldata$`Sample name`]
ddsTxi <- DESeqDataSetFromTximport(txi,
                                   colData = coldata, 
                                   design = ~ Treatment)
```




### removing rows with low gene counts and keeping rows that have at least 10 reads total

`` helps reduce the size of the object which increases the computational speed``


```{r}
keep <- rowSums(counts(ddsTxi)) >= 10
ddsTxi <- ddsTxi[keep, ]
```

 running DESeq
 
if needed collapsing the technical replicates``


```{r}
ddsTxi <- DESeq(ddsTxi)
```

saving the results``


```{r}
results_ddsTxi <- results(ddsTxi)
results_ddsTxi
```

 Exploring results


```{r}
summary(results_ddsTxi)
```

contrasts


```{r}
resultsNames(ddsTxi)
```

 MA plot


```{r}
plotMA(results_ddsTxi)
```

ordering results

ordering the results to see the 10 genes with the most significant p value``


```{r}
resOrdered <- results_ddsTxi[order(results_ddsTxi$pvalue),]
head(resOrdered,n=10)
```

reqired libraries

```{r}
BiocManager::install("biomaRt")
BiocManager::install("annotables")
BiocManager::install("org.Hs.eg.db")
BiocManager::install("EnsDb.Hsapiens.v86")
library("biomaRt")
library("annotables")
library("org.Hs.eg.db")
library("EnsDb.Hsapiens.v86")
library("tidyverse")
```

input list of Enselbl IDs

```{r}
ensembl.ids <- read.delim("D:Ensembl IDss.txt", header = F)
```

 translating with biomart

seeing the available databeses and their values``


```{r}
listEnsembl()
```

choosing the database``


```{r}
ensembl <- useEnsembl(biomart = "genes")
```

seeing which dataset to use``


```{r}
datasets <- listDatasets(ensembl)
```

connecting the dataset and databese we want to use``


```{r}
ensembl.conection <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
```

seeing which attributes we need``
```{r}
attributes <- listAttributes(ensembl.conection)
```



seeing which filter we need``


```{r}
filters <- listFilters(ensembl.conection)
```

buildung the query``



```{r}
getBM(attributes = c("ensembl_gene_id","external_gene_name"),
      filters = "ensembl_gene_id",
      values = ensembl.ids$V1,
      mart = ensembl.conection)
```

###GSVA

```{r}
BiocManager::install("GSVA")
library("GSVA")

```
loading of gene sets from GMT file
```{r}
gene_sets <- read.delim("D:h.all.v2023.1.Hs.symbols.gmt",
sep = "\t", header = FALSE, stringsAsFactors = FALSE)
```


```{r}
BiocManager::install("shinydashboard")
BiocManager::install("shinyjs")
BiocManager::install("limma")
library("shinydasboard")
```


```{r}
res <- igsva()
```

